---
title: "Data Analysis"
author: "Luisa Pires"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---


This markdown shows an example of an analysis using the daily country data on number of deaths, recovered and confirmed cases of COVID 19. 

Here we focus on Portugal, Spain, UK, Germany and Italy. Analysis performed:

1. Daily differences (absolute,  percentage and 7-day rolling average)
2. % differences in accumulated data 
3. indicators per million inhabitants
4. Analysis of reference points (days since lock down, days since 50th death reported)


Data from John Hopkins with thanks:

1. https://github.com/pomber/covid19
2. https://github.com/CSSEGISandData/COVID-19

Description of the data from [2]: "This is the data repository for the 2019 Novel Coronavirus Visual Dashboard operated by the Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE). Also, Supported by ESRI Living Atlas Team and the Johns Hopkins University Applied Physics Lab (JHU APL)."


```{r setup, echo=FALSE,  cache=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)

knitr::opts_knit$set(width = 75)

```

# Setup

Necessary libraries for running the analysis.

```{r}

library(knitr)
library(prettydoc)
library(tidyverse)
library(httr)
library(jsonlite)
library(plotly)
library(kableExtra)
library(zoo)

```

# Auxiliary functions

Here are some auxiliary functions to display the data in a consistent way.
```{r, echo=TRUE}

# compute variable per million inhabitants
# usage:
# df %>%  per_million_inhab(.,var_columns=c("deaths","recovered"))
per_million_inhab = function(df, var_columns){
  
df = df %>%
    mutate_at(.vars = vars(var_columns), 
              .funs = funs(round(./population, 2)))

}

# compute variable in a 7-day rolling average basis
# usage:
# df %>%  rolling_aver_7(.,var_columns=c("deaths","recovered"))
rolling_aver_7 = function(df, var_columns){
df = df %>%
    mutate_at(.vars = vars(var_columns), 
              .funs = funs(round(
                rollmean(., 7, align="right",na.pad = T),
                2)))

}

# compute the variable's percentage increase
# usage:
# df %>%  perc_increase(.,var_columns=c("deaths","recovered"))
perc_increase = function(df, var_columns){
df = df %>%
    mutate_at(.vars = vars(var_columns), 
              .funs = funs(round(
                (.-lag(., n = 1, default = NA))/lag(., n = 1, default = NA),
                2)*100))

}


# add all the transformations together
# usage:
# df %>%  transform_data(.,var_columns=c("deaths","recovered"), per_million = T)
transform_data = function (df, var_columns,  per_million = F, perc_inc = F, rolling = F ){

  if(per_million) df  = df %>%  per_million_inhab(.,var_columns=var_columns)
  if(rolling) df  = df %>%  rolling_aver_7(.,var_columns=var_columns)
  if(perc_inc) df  = df %>%  perc_increase(.,var_columns=var_columns)

  return(df) 
}


```


```{r, echo = F}

# adust axix 
adjust_axis = function(df, xvariable, last_n_days = 30){
  
  if (xvariable == "date")  {
   
     df = dplyr::filter(df, date > (Sys.Date() - last_n_days))
  }
  else {
  
      df = filter(df, get(xvariable) > 0)
  }
  
return(df)
  }

# make a pretty comparative table
make_table_comparison = function(df, xvariable, yvariable, countries, last_n_days, ...) {
 
  
   df = select(df, one_of(yvariable, xvariable, "country","population")) %>% 
     transform_data(var_columns = yvariable, ...) %>% 
     adjust_axis(xvariable, last_n_days) %>% 
     select(-population)
  
  df = df %>% 
    pivot_wider(names_from = "country", values_from = yvariable) %>%
    mutate_at(vars(countries), function(x) {
      cell_spec(x, bold = T,
                color = spec_color(x, end = 0.9))
    }) %>%
    
    kable(escape = F, align = "c") %>%
    kable_styling(c("striped", "hover"))
  
  return(df)
  
}

# make a pretty table
make_table = function(df, last_n_days = 10) {
 
   df %>%  
    tail(last_n_days) %>%
    kable() %>%
    kable_styling()
  
}

# make a standard bar chart with parameterized axes
chart_variable_bar = function(df, xvariable, yvariable, last_n_days = 30, ...) {
 
  #  adjust the data according to the variable 
   df = df %>% 
     transform_data(var_columns = yvariable) %>% 
     adjust_axis(xvariable, last_n_days)
  
  p = ggplot(data = df, aes_string(x = xvariable, y = yvariable, fill =
                                     "country")) +
    geom_bar(stat = "identity", position = position_dodge()) +
    scale_fill_brewer(palette = "Paired") +
    theme_minimal() + 
    xlab(xvariable) + 
    ylab(yvariable) 

  
  ggplotly(p)
  
}

# make a standard bar chart with parameterized axes
chart_variable_facets = function(df, xvariable, yvariable, last_n_days = 30, ...) {
 
  #  adjust the data according to the variable 
   df = df %>% 
     transform_data(var_columns = yvariable, ...) %>% 
     adjust_axis(xvariable, last_n_days)
  
  p = ggplot(data = df, aes_string(x = xvariable, y = yvariable, fill =
                                     "country"))  +
    geom_bar(stat = "identity", position = position_dodge()) +
    scale_fill_brewer(palette = "Paired") +
    theme_minimal() + 
    xlab(xvariable) + 
    ylab(yvariable) +
      facet_grid(cols=vars(country)) + 
    theme(axis.text.x=element_text(size=rel(1), angle=45))
  
  ggplotly(p)
  
}

# make a standard line chart with parameterized axes
chart_variable_line = function(df, xvariable, yvariable, last_n_days = 10,  ...) {
  
    #  adjust the data according to the variable 
   df = df %>% 
     transform_data(var_columns = yvariable, ...) %>% 
     adjust_axis(xvariable, last_n_days)
 
  p = ggplot(data = df,
             aes_string(x = xvariable, y = yvariable, colour = "country")) +
    geom_line() +
    scale_color_brewer(palette = "Paired") +
    theme_minimal() + 
    xlab(xvariable) + 
    ylab(yvariable)
  
 ggplotly(p)
 
}


# find peaks in time series 
find_peaks = function (x, thresh = 0)
{
  pks =
    which(diff(sign(diff(x, na.pad = FALSE)), na.pad = FALSE) < 0) + 2
  if (!missing(thresh)) {
    pks[x[pks - 1] - x[pks] > thresh]
  }
  else
    pks
}

# find dominat frequencies
find_freq = function(x, plot = F) {
  x[is.na(x)] = 0
  
  del = 1 # sampling interval
  x.spec = spectrum(x, log = "no", plot = FALSE)
  spx = x.spec$freq / del
  spy = 2 * x.spec$spec
 if(plot) {plot(spy ~ spx,
       xlab = "frequency",
       ylab = "spectral density",
       type = "l")}
  
  peaks = find_peaks(spy)[1:3]
  peak_values = sort(spy[peaks], index.return = TRUE)
  freq = spx[peaks[peak_values$ix]]
  freq = freq[freq>0.1]
  period = round(1 / freq[length(freq)], 2)
  
  return(period)
}

```

# Get & prepare the data

Here we get the data from  "https://pomber.github.io/covid19/timeseries.json". A few inconsistencies have been detected from worldmeter.

1. United Kingdom (16-19 April)
2. Spain (last 16019 April)

```{r}

#path to the API and request data
path = "https://pomber.github.io/covid19/timeseries.json"
request = GET(url = path)

# get the response
response = content(request, as = "text", encoding = "UTF-8")

# data from json
data_json = fromJSON(response) %>%
  data.frame()

# show data
data_json %>% 
  select(1:6) %>%
  make_table(5)

```

Prepare the data

Here we do the following transformations:

1. Split the first column into dimensions
2. Dimensions are further split into confirmed, deaths and recovered

```{r}

# find the last "."
data = enframe(unlist(data_json)) %>%
  mutate(position = str_locate_all(name, "\\.")) %>%
  rowwise() %>%
  mutate(position = max(position)) %>% 
  ungroup()

# replace last dot with space
str_sub(data$name, data$position, data$position) = " "

# split into countries, dimensions and sub-dimensions
data = data %>%
  separate(.,
           name,
           c("country", "dimension"),
           sep = " " ,
           remove = FALSE) %>%
  mutate(dimension1 = (str_extract(dimension, "[aA-zZ]+")),
         datapointid = (str_extract(dimension, "[0-9]+"))) %>%
  select(-name, -dimension, -position) %>%
  pivot_wider(names_from = dimension1,
              values_from = value) %>%
  select(-datapointid)

# show data
data %>% 
  make_table(10)

```
Now we change formats to ensure:

1. dates are formatted as such
2. sub-dimensions are numeric

```{r}

data_clean = data %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d")) %>%
  mutate_at(vars(confirmed, deaths, recovered), as.numeric)

```

We can now compute daily numbers for the countries.

```{r}

data_clean = group_by(data_clean, country) %>%
  mutate(
    previous_deaths  = lag(deaths, n = 1, default = NA),
    previous_confirmed  = lag(confirmed, n = 1, default = NA),
    previous_recovered  = lag(recovered, n = 1, default = NA),
    daily_deaths  = deaths - previous_deaths,
    daily_confirmed  = confirmed - previous_confirmed,
    daily_recovered  = recovered - previous_recovered ) %>% 
  select(-starts_with("previous"))

data_clean %>% 
  select (country, date, starts_with("daily")) %>% 
  make_table(10)

```

# User-defined parameters

The user can now select the countries of interest and add extra information. For the countries that are of interest we need to add the population (in millions) and the day of lockdown.

```{r}

#countries of interest
country = c("Portugal", "Spain", "Italy", "United.Kingdom", "Germany")

#  population of countries
population = c(10.28, 46.94, 60.36, 66.65 , 83.02)

# lockdown start days 
lockdown =
  as.Date(c(
    "2020-03-18",
    "2020-03-14",
    "2020-03-09",
    "2020-03-23",
    "2020-03-22"
  ), format = "%Y-%m-%d")

# make data frame with reference data
interest_countries = data.frame(country, population, lockdown, stringsAsFactors = F)

# join the daily data with reference data
data_clean = data_clean %>%
  inner_join(interest_countries, by="country")

```


# Analysis

In order to better compare countries and measures, we add a few reference columns such as *days since lock-down* and *days since 50th death reported*.

## Adding reference dates

We now add a few reference points.

### Days since lock-downs 

In order to compare the effect of lock-down, we add a flag that counts the number of days since lock-down for these countries

```{r}


# count lockdiwn days
data_clean = data_clean %>%
  mutate(days_since_lockdown = ifelse( date >= lockdown,
                                      1,
                                      0
  )) %>%
  mutate(days_since_lockdown = cumsum(days_since_lockdown))

# show data
data_clean %>% 
  select(country,date, days_since_lockdown) %>%  
  make_table(10)

```

###  Days since 50 deaths reported

We now create a flag for the days since the 50th death reported.

```{r}

# find 50th death
data_clean = data_clean %>%
  mutate(days_after_50_death = ifelse(deaths >= 50, 1, 0)) %>%
  mutate(days_after_50_death = cumsum(days_after_50_death))

# show table
data_clean %>% 
  select(country, date,days_after_50_death) %>% 
  make_table(10)

```

# Comparative analysis

We can do now comparative analysis between countries. The folowing flags help define the anaysis:

* *per_million = T* indicates the y-variable is divided by the country's population (in_millions)
* *per_increase = T* indicates the y-variable is measured in percentage increase (%) from the day before
* *rolling = T* ndicates the y-variable is computed using a 7-day rolling average

## Standard questions

* What is the daily increase for confirmed cases and deaths?

```{r}

variables = c("daily_confirmed","daily_deaths")

data_clean %>% transform_data(var_columns = variables, perc_inc = T) %>% 
  select(date, country, one_of(variables)) %>% 
  filter(date == max(date)) %>% 
  make_table()
  

```

* What is the trend curve for the daily confirmed cases?

```{r}

chart_variable_facets(df = data_clean, 
                    xvariable = "date", 
                    yvariable = "daily_confirmed",
                    rolling = T,
                    last_n_days = 30)

```
The daily increase in numbers is below:

```{r}

data_clean %>% 
  make_table_comparison(xvariable  = "date",
                        yvariable  = "daily_confirmed", 
                        countries = interest_countries$country,
                        perc_inc = T,
                        rolling = T,
                        last_n_days = 2)

```

* What is the trend curve for the daily deaths cases?

```{r}

chart_variable_facets(df = data_clean, 
                    xvariable = "date", 
                    yvariable = "daily_deaths",
                    rolling = T,
                    last_n_days = 30)

```

The daily increase in numbers is below:

```{r}

data_clean %>% 
  make_table_comparison(xvariable  = "date",
                        yvariable  = "daily_deaths", 
                        countries = interest_countries$country,
                        perc_inc = T,
                        rolling = T,
                        last_n_days = 2)

```

* How do countries compare regarding number of deaths since 50th death reported?

```{r}

chart_variable_line(df = data_clean, 
                    xvariable = "days_after_50_death", 
                    yvariable = "deaths", 
                    rolling = T)

```


* How do countries compare regarding number of deaths since 50th death reported taking into acccount the population?

```{r}

chart_variable_line(df = data_clean, 
                    xvariable = "days_after_50_death",
                    yvariable =  "deaths", 
                    rolling = T, 
                    per_million =T)

```



## Lock-down affect

* Haw the lockdown influence the  % increase in daily confirmed cases?

```{r}

chart_variable_line(df = data_clean,
                    xvariable = "days_since_lockdown", 
                    yvariable = "daily_confirmed", 
                    perc_inc = T, 
                    rolling = T)

```

The table version of this chart is below:

```{r}

data_clean %>% 
  make_table_comparison(xvariable  = "days_since_lockdown",
                        yvariable  = "daily_confirmed", 
                        countries = interest_countries$country,
                        perc_inc = T,
                        rolling = T)

```

* Haw has the lockdown influence the trend in daily deaths?

```{r}

chart_variable_line(df = data_clean, 
                    xvariable = "days_since_lockdown",
                    yvariable = "daily_deaths",  
                    rolling  = T, 
                    per_million = T)

```



* Rolling average of daily difference in confirmed cases 

```{r}

chart_variable_line(df = data_clean, 
                    xvariable = "days_since_lockdown", 
                    yvariable = "daily_confirmed",
                    rolling = T, 
                    per_million = T)

```

## Frequencies

In Spain and Italy the numbers show a periodic trend on confirmed cases are noticeable. Below we compute that period.


```{r}


data_clean %>%
  group_modify(~ {
     find_freq(.x$daily_confirmed, plot=F) %>%
     tibble::enframe(name = "aux", value = "period (days)")
  }) %>% select(-aux) %>%
  make_table()

```

We can do the same for the  daily deaths. 

```{r}

data_clean %>%
  group_modify(~ {
     find_freq(.x$daily_deaths, plot=F) %>%
     tibble::enframe(name = "aux", value = "period (days)")
  }) %>% select(-aux) %>%
  make_table()
```

